Inferring stellar rotation periods using Gaussian processes

### 2016-12-07 documents/ms.tex
outline caveat, equation typo, anti-correlation description

@@

\subsection{Kernel function choice and interpretation of hyper-parameters}
The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos^2\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
It may in fact be even {\it more} suited to modelling stellar
light curves as it describes a Gaussian in frequency space.
It is easy to imagine a differentially rotating star with a period that is a
Gaussian in frequency space: the mean frequency would be the frequency at the
most active latitude, at or near which spots spend the majority of their time
and the tails would be occupied by spots that drift near the equator or poles.
The main difference between this cosine and the QP function is that the cosine
function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.

>>>>

At this point it is important to highlight the following caveat.
Our method has been tuned to reproduce the rotation periods of the
\citet{Aigrain2015} simulated light curves.
These light curves were simulated using a very good but still imperfect model.
The input parameters and their distributions were chosen to be as realistic as
possible given current observations.
Unfortunately however, there are limited data on some of these parameters for
stars other than the Sun: for example the distrubution of spot lifetimes or
activity cycle length.
Ensuring that we can recover the rotation periods of these simulated light
curves does not necessarily ensure that we can recover the rotation periods of
real \kepler\ light curves with equal accuracy and precision.
Even using priors on the QP kernel parameters that have been derived from the
\citet{Aigrain2015} sample may be dangerous.
This caveat should be taken into consideration when this method is applied to
real data.

@@

\subsection{Kernel function choice and interpretation of hyper-parameters}
The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
It may in fact be even {\it more} suited to modelling stellar
light curves as it describes a Gaussian in frequency space.
It is easy to imagine a differentially rotating star with a period that is a
Gaussian in frequency space: the mean frequency would be the frequency at the
most active latitude, at or near which spots spend the majority of their time
and the tails would be occupied by spots that drift near the equator or poles.
The main difference between this cosine and the QP function is that the cosine
function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
In other words, we do not want to force anti-correlation of points that are
$\frac{1}{2}$ a period apart.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.
